---
title: "Auto Claims Analysis"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```




# Introduction

I worked with a dataset of auto claims filed by customers of an automobile insurance company located in the southwest and western regions of the United States.

Insurance companies depend on accurate pricing estimates to maintain profitability. Auto policies must be priced so that the insurance company makes a profit in the long run, given the costs of their customers' payouts for accident repairs, total loss car replacements, medical fees, and legal fees.

The executives at this insurance company have noticed declining profitability over the last several years and are in neeed of data science consultant to evaluate their claims data and make recommendations on pricing, customer behavior, and car insurance policy adjustments.

The objective of this project is to perform an exploratory data analysis on the claims_df dataset and produce an executive summary of your key insights and recommendations to the executive team at the insurance company.

Before you begin, take a moment to read through the following insurance company terms to familiarize yourself with the industry:  [Auto Insurance
Terms](https://www.iii.org/article/auto-insurance-jargon-buster "Auto Insurance Terms"){target="_blank"} 



## Auto Claims Data

The `claims_df` data frame is loaded below and consists of 6,249 auto claims submitted by customers of the insurance company. The rows in this data frame represent a single claim with all of the associated features that are displayed in the table below.




```{r}
library(tidyverse)

claims_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/claims_df.rds'))

```


### Raw Data

```{r}

claims_df

```
<br>
<br>


# Exploratory Data Analysis



The goal of your analysis should be discovering which variables drive the differences between customers with large lifetime values and customers who cost the company more than they provide in revenue through monthly premiums.

The following questions were designed and explored based on the available data. 



## Question 1
Are there types of customers, based on their domiciled state, that are highly profitable?
```{r Question 1}
library(ggplot2)
library(dplyr)
library(tidyverse)

A <- ggplot(data=claims_df,aes(x=reorder(customer_state, customer_lifetime_value), y=customer_lifetime_value,fill = customer_state)) + geom_bar(stat="identity",show.legend = FALSE) + xlab("State") + ylab("Customer Lifetime Value")+expand_limits(y=-1000000) +ggtitle("Customer State verses Customer life time value") + theme(plot.title=element_text(hjust=0.5))
options(scipen = 999)
plot(A)

center_title <- theme(plot.title=element_text(hjust=0.5))

Total_California_Revenue <- claims_df %>% filter(customer_state == "California") %>% summarize(sum(customer_lifetime_value))

Total_Oregon_Revenue <- claims_df %>% filter(customer_state == "Oregon") %>% summarize(sum(customer_lifetime_value))

Total_Arizona_Revenue <- claims_df %>% filter(customer_state == "Arizona") %>% summarize(sum(customer_lifetime_value))

Total_Washington_Revenue <- claims_df %>% filter(customer_state == "Washington") %>% summarize(sum(customer_lifetime_value))

Total_Nevada_Revenue <- claims_df %>% filter(customer_state == "Nevada") %>% summarize(sum(customer_lifetime_value))

df0 <- data.frame(Total_California_Revenue,Total_Oregon_Revenue,Total_Arizona_Revenue,Total_Washington_Revenue,Total_Nevada_Revenue)

colnames(df0) <- c("Total California Revenue","Total Oregon Revenue","Total Arizona Revenue","Total Washington Revenue","Total Nevada Revenue")

view(df0)
```

The graph clearly indicates that the customers based out of the states of California,Oregon and Arizona provided very profitable customer lifetime value earning the company net revenues of $1898706 , $1670078 and $1129186 respectively. The customers from the states of Washington and Nevada showcased the lowest return on investment generating $480355 and $583650 respectively in net income. 



```{r}




```



## Question 2
Do certain policies have a lower number of claims, leading to large profits?
**Answer**:

```{r Question 2}
library(ggplot2)
library(dplyr)
library(tidyverse)

B <- ggplot(data=claims_df,aes(x=policy, y=total_claims,fill=policy)) + geom_bar(stat="identity",show.legend = FALSE) + xlab("Customer's Policy") + ylab("Total Claims") + ggtitle("Customer's Insurance Policy verses Customer's \n Total Numer of Claims") + theme(plot.title=element_text(hjust=0.5))
plot(B)

Total_claims<- claims_df %>% summarize(sum(total_claims))
view(Total_claims)
Total_Corporate_Claims <- claims_df %>% filter(policy == "Corporate") %>% summarize(sum(total_claims))

Total_Personal_Claims <- claims_df %>% filter(policy == "Personal") %>% summarize(sum(total_claims))

Total_Special_Claims <- claims_df %>% filter(policy == "Special") %>% summarize(sum(total_claims))

df1 <- data.frame(Total_claims,Total_Corporate_Claims,Total_Personal_Claims,Total_Special_Claims)

colnames(df1) <- c("Total Claims","Total Corporate Claims","Total Personal Claims","Total Special Claims")

view(df1)
```
There were a total of 14946 claims filed. As per the graph and queries  performed on the data, the customers on the Personal Policy had the highest number of claims (11131) followed by the customers on the Corporate policy (3175). The customers on the Special policy filed the least number of claims (640). 


## Question 3
Does "Sales channel" utilized to on-board customers have an impact on the number of claims?
**Answer**:


```{r Question 3}
library(ggplot2)
library(dplyr)
library(tidyverse)

C <- ggplot(data=claims_df,aes(x=sales_channel,fill=sales_channel)) + geom_histogram(stat="count",show.legend = FALSE) +ggtitle("Total Customers brought in by each of the Sales Channels") + center_title +xlab("Sales Channel")+ylab("Number of Customers Onboarded")
plot(C)

Agent_attracted_customer <- claims_df %>% filter(sales_channel == "Agent")
str(Agent_attracted_customer)
view(Agent_attracted_customer)
Total_Agent_Claims <- claims_df %>% filter(sales_channel == "Agent") %>% summarize(sum(total_claims))

view(Total_Agent_Claims)

Branch_attracted_customer <- claims_df %>% filter(sales_channel == "Branch")
str(Branch_attracted_customer)
view(Branch_attracted_customer)
Total_Branch_Claims <- claims_df %>% filter(sales_channel == "Branch") %>% summarize(sum(total_claims))

view(Total_Branch_Claims)

Call_attracted_customer <- claims_df %>% filter(sales_channel == "Call Center")
str(Call_attracted_customer)
view(Call_attracted_customer)
Total_Call_Claims <- claims_df %>% filter(sales_channel == "Call Center") %>% summarize(sum(total_claims))

view(Total_Call_Claims)

Web_attracted_customer <- claims_df %>% filter(sales_channel == "Web")
str(Web_attracted_customer)
view(Web_attracted_customer)
Total_Web_Claims <- claims_df %>% filter(sales_channel == "Web") %>% summarize(sum(total_claims))

view(Total_Web_Claims)

df2 <- data.frame(Total_Agent_Claims/2359,Total_Branch_Claims/1771,Total_Call_Claims/1218,Total_Web_Claims/901)

colnames(df2) <- c("Agent Attracted Customer claims (Average)","Branch Attracted Customer claims (Average)","Call Center Attracted Customer claims (Average)","Web Attracted Customer claims (Average)")

view(df2)
```
The histogram displays the distribution of the number of customers acquired using each of the various channels of marketing. An in-depth analysis was performed on the filtered data on each of the marketing channels in order to determine average claims filed by customers on-boarded by the particular channel of sales. The data suggests that the Agent attracted customers had the lowest claims on average (2.38 per customer) while the customers who signed up using the website had the highest claims on average (2.41 per customer). 



## Question 4
**Question**:
Is there an ideal subscription length (months) which ensures maximum revenue for the company?

**Answer**:

```{r Question 4}
summary(claims_df)

Good_customers <- claims_df %>% filter(customer_lifetime_value > 922.1)
view(Good_customers)

summary(Good_customers)

D <- ggplot(Good_customers, aes(months_policy_active)) + geom_density() + xlab("Number of months since the policy was active") + ylab("Density")+ ggtitle(" Distribution of the Months Policy was active for the company's \n highly profitable clients") + center_title
plot(D)


```
A new data set was formed consisting of all the customers which yielded net revenues of more than the average revenue per customer (> $922.1). A density plot was formed based on the updated data set. The graph indicates that the customers with active policies in between 40 to 55 months turned out to be highly profitable in terms of revenue generated. The company's highly profitable clients had their policies active for an average of 46.42 months.



## Question 5

**Question**:
How does the class of vehicle impact the customer life time revenue generated for the company?


**Answer**:

```{r Question 5}

E <- ggplot(claims_df, aes(vehicle_class, customer_lifetime_value)) + geom_boxplot() +xlab("Vehicle Class") +ylab("Customer Lifetime Value") + ggtitle(" The impact of Customer's Vehicle Class on the \n Customer Lifetime Value") +center_title + coord_flip()
plot(E)

twodoorcar_lifetime_value <- claims_df %>% filter(vehicle_class == "Two-Door Car") %>% summarize(mean(customer_lifetime_value))

fourdoorcar_lifetime_value <- claims_df %>% filter(vehicle_class == "Four-Door Car") %>% summarize(mean(customer_lifetime_value))

sportscar_lifetime_value <- claims_df %>% filter(vehicle_class == "Sports Car") %>% summarize(mean(customer_lifetime_value))

suv_lifetime_value <- claims_df %>% filter(vehicle_class == "SUV") %>% summarize(mean(customer_lifetime_value))

luxurycar_lifetime_value <- claims_df %>% filter(vehicle_class == "Luxury Car") %>% summarize(mean(customer_lifetime_value))

luxurysuv_lifetime_value <- claims_df %>% filter(vehicle_class == "Luxury SUV") %>% summarize(mean(customer_lifetime_value))

df3 <- data.frame(twodoorcar_lifetime_value,fourdoorcar_lifetime_value,sportscar_lifetime_value,suv_lifetime_value,luxurycar_lifetime_value,luxurysuv_lifetime_value)

colnames(df3) <- c("Two Door Car Lifetime Revenue (Average)","Four Door Car Lifetime Revenue (Average)","Sports Car Lifetime Revenue (Average)","SUV Lifetime Revenue (Average)","Luxury Car Lifetime Revenue (Average)","Luxury SUV Lifetime Revenue (Average)")

view(df3)

```
A box-plot was created in order to understand which of the vehicle class generated the highest revenue for the company. The box plot revealed that the Luxury Suvs, Luxury Cars, and SUVs generated highest average lifetime revenues of $6381, $5669, and $1861 respectively. However, Two door and Four door cars fell on the low end of the spectrum yieldings average lifetime revenues of $269 and $271 respectively. 


## Question 6

**Question**:
What kind of coverage had the greatest positive impact on the company's revenue generation?  


**Answer**:


```{r Question 6}
F <- ggplot(claims_df, aes(coverage, customer_lifetime_value,fill=coverage)) + geom_violin(show.legend = FALSE) +xlab("Coverage") +ylab("Customer Lifetime Value") + ggtitle(" Net Revenue brought in by each of the Insurance \n Coverage Types") + center_title
plot(F)

basic_lifetime_value <- claims_df %>% filter(coverage == "Basic") %>% summarize(mean(customer_lifetime_value))

extended_lifetime_value <- claims_df %>% filter(coverage == "Extended") %>% summarize(mean(customer_lifetime_value))

premium_lifetime_value <- claims_df %>% filter(coverage == "Premium") %>% summarize(mean(customer_lifetime_value))

df4 <- data.frame(basic_lifetime_value,extended_lifetime_value,premium_lifetime_value)

colnames(df4) <- c("Basic Policy Lifetime Revenue (Average)","Extended Policy Lifetime Revenue (Average)","Premium Policy Lifetime Revenue (Average)")

view(df4)




```
A Violin Plot was plotted on the original data set. The Violin Plot helps understand the distribution of each coverage type on the revenue scale. The graph suggests that customers on the Extended and Premium coverage have most of the data points above the zero mark (below which the company starts incurring losses). On the contrary, the Lifetime revenues for the Basic coverage are evenly distributed around the value of zero indicating a great number of customers on the coverage generated net revenues of less than zero for the company. The average Customer Lifetime Value of customers on the Basic, Extended and Premium coverage came out to be $326, $1538 and, $2879 respectively. 

## Question 7

**Question**:
How do Customer Gender and Residential Demography influence company's profitability?


**Answer**:

```{r Question 7}
summary(claims_df)

E <- ggplot(data=Good_customers,aes(x= residence_type, y = customer_lifetime_value,color=gender,fill=gender))  +geom_bar(stat="identity",show.legend = FALSE) + facet_grid( ~gender) + xlab("Customer's Residence Type") + ylab("Net Revenue Generated") +ggtitle("The correlation of Customer's Gender and Residence Type with \n the Lifetime Revenue Generated") + center_title
plot(E)


suburban_women <- Good_customers %>% filter(residence_type == "Suburban",gender == "Female") %>% summarize(sum(customer_lifetime_value))

urban_women <- Good_customers %>% filter(residence_type == "Urban",gender == "Female") %>% summarize(sum(customer_lifetime_value))

Rural_women <- Good_customers %>% filter(residence_type == "Rural",gender == "Female") %>% summarize(sum(customer_lifetime_value))

suburban_men <- Good_customers %>% filter(residence_type == "Suburban",gender == "Male") %>% summarize(sum(customer_lifetime_value))

urban_men <- Good_customers %>% filter(residence_type == "Urban",gender == "Male") %>% summarize(sum(customer_lifetime_value))

Rural_men <- Good_customers %>% filter(residence_type == "Rural",gender == "Male") %>% summarize(sum(customer_lifetime_value))



suburban_women_count <- Good_customers %>% filter(residence_type == "Suburban",gender == "Female") %>% count()

urban_women_count <- Good_customers %>% filter(residence_type == "Urban",gender == "Female") %>% count()

Rural_women_count <- Good_customers %>% filter(residence_type == "Rural",gender == "Female") %>% count()

suburban_men_count <- Good_customers %>% filter(residence_type == "Suburban",gender == "Male") %>% count()

urban_men_count <- Good_customers %>% filter(residence_type == "Urban",gender == "Male") %>% count()

Rural_men_count <- Good_customers %>% filter(residence_type == "Rural",gender == "Male") %>% count()

Net_revenue_good_customer <- data.frame(suburban_women,urban_women,Rural_women,suburban_men,urban_men,Rural_men)

Suburban_women_c <- c(suburban_women , suburban_women_count)
view(Suburban_women_c)
Suburban_men_c <- c(suburban_men,suburban_men_count)
view(Suburban_men_c)
Urban_women_c <- c(urban_women,urban_women_count)
Urban_men_c <- c(urban_men,urban_men_count)

Rural_women_c <- c(Rural_women,Rural_women_count)
Rural_men_c <- c(Rural_men,Rural_men_count)

Final_df <- data.frame(Suburban_women_c,Suburban_men_c,Urban_women_c,Urban_men_c,Rural_women_c,Rural_men_c)

Demography <- c("Suburban","Urban","Rural","Suburban","Urban","Rural")
Gender <- c("Female","Female","Female","Male","Male","Male")
Net_Revenue <- c(2531073,2011879,731172,552967,815521,466730)
Counts <- c(931,692,288,219,315,191)

f_df <- data.frame(Demography,Gender,Net_Revenue,Counts)
view(f_df)

F <- ggplot(f_df,aes(x= Demography, y = Counts,color=Gender,fill=Gender))  +geom_bar(stat="identity",show.legend = FALSE) + facet_grid( ~Gender) + xlab("Customer's Residence Type") + ylab("Count") +ggtitle("The Strength of Customer's Gender and Residence Type on the \n Lifetime Revenue Generated Scale") + center_title
plot(F)


```
The previously created subset of customers netting above the mean revenue value was used to run an analysis. Two bar graphs were plotted. The first graph presents an overview of Net revenue generated by each customer based on their residential locality and gender. Suburban Women grossed the company $2531073 in profits which is way higher than Men from the same locality ($552967). Upon further investigation of the data, another bar chart was plotted to explore the strength and distributions of the company's highly profitable customers. The graph revealed that there were 931 Suburban Women who generated above average revenues for the company which is significantly higher than the males from the same locality (219). Both the graphs along with the Data Frames communicate that of all of the company's highly profitable customers, Women tend to be greater in number.     



## Question 8



**Question**:
How does customer Gender and Monthly Premium impact company revenues?

**Answer**:


```{r Question 8}

G <- ggplot(data=claims_df,aes(x= monthly_premium, y = customer_lifetime_value,color=gender))  +geom_point(show.legend = FALSE) + facet_grid( ~gender) +xlab("Monthly Premium") +ylab("Customer Lifetime Value") +ggtitle("The impact of Customer Gender and Monthly Premium Amount on \n the Net Revenue Generated") + center_title
plot(G)

Male_losses <- claims_df %>% filter(gender == "Male",customer_lifetime_value < 0) %>% summarize(sum(customer_lifetime_value))

Female_losses <- claims_df %>% filter(gender == "Female",customer_lifetime_value < 0) %>% summarize(sum(customer_lifetime_value))

Losses <- data.frame(Male_losses,Female_losses)

colnames(Losses) <- c("Net Male Losses","Net Female Losses")
view(Losses)
```
A scatter plot was created to understand how the customer gender and monthly premium impact company's revenues. The graph indicates that the more a customer pays in the monthly premiums, the greater the chance that he/she will earn revenue for the company. The patterns also revealed that Male customers, on average, incurred the company more losses in comparison with the Female customers. The company had to pay a grand total of $1213122 to their Male clients in claims while they had to spend a far lesser amount of $758704 covering up their Female customer claims. 

```{r extra work}

```

# Summary of Results

## Executive Summary

Introduction

The goal of the project was to perform an explanatory data analysis on data set of the auto claims filed by customers of the automobile insurance company in order to identify the factors impacting the profitability of the company. The prime objective of the analysis was to look into the segments of the company’s most and the least profitable customers for any clear patterns and indications which could potentially help the company minimize losses which in turn would increase the revenues. A set of 8 questions was designed and answered so as to maximize the value of the insights extracted. The questions explored how customer demography, channel of sales used to onboard the customer, customer’s class of vehicle and customer gender impacted the revenue generated. The questions also investigated the influence of insurance policy and coverage type on the net revenue of the company.

Highlights and Key Findings

A lot of interesting observations were made while analysing the data set. The customers based out of the states of California, Oregon and Arizona proved to be the most profitable of all the other states. Knowing this could help the company run targeted marketing campaigns in the areas where customer base generates maximum revenues for the company. The customers on the Personal Policy had the highest number of claims. This insight could help the company decide which policy to penalize or incentivize in order to maximize the revenues. The Agent attracted customers had the lowest claims on average (2.38 per customer) while the customers who signed up using the website had the highest claims on average (2.41 per customer). This knowledge can be of great value to the company’s marketing and sales team when choosing which sales channel to exhaust the most. The customers with active policies in between 40 to 55 months turned out to be highly profitable in terms of revenue generated. The company's highly profitable clients (the customers who yielded net revenues of more than the average revenue per customer (> $922.1)) had their policies active for an average of 46.42 months. This awareness could encourage the company to launch targeted advertisement campaigns to attract long term clients. The Luxury Suvs, Luxury Cars, and SUVs generated highest average lifetime revenues for the company. Knowing which type of vehicle earns the company the most money could help the decision makers make informed decisions. The customers on the Extended and Premium coverage have most of the data points above the zero mark (below which the company starts incurring losses). On the contrary, the Lifetime revenues for the Basic coverage are evenly distributed around the value of zero indicating a great number of customers on the coverage generated net revenues of less than zero for the company. This information could prove to be of great value when designing the company’s loss mitigation plan. Suburban Women grossed the company $2,531,073 in profits which is way higher than that of any other demographic or gender. This discovery could help the company run highly tailored offers to attract more clients from the same demography. The company had to pay a grand total of $1,213,122 to their Male clients in claims while they had to spend a far lesser amount of $758,704 covering up their Female Customer claims. These stats could help the company set insurance premiums in order to minimize the losses. 

Recommendations

The clientele from the states of California and Oregon made the company $189,870,6 and $167,007,8 in revenues respectively which is significantly higher than that of any other state. Based on this observation, the company should run massive advertisement campaigns targeting these two states. This would enhance the likelihood of company acquiring customers that would prove highly profitable.

A total of 11131 claims were filed by the customers holding the Personal insurance policy while significantly lower number of claims were received for customers on the Corporate (3175 claims) and Special plans (640). Considering the stats, the company should look to run B2B marketing campaigns and offer Employer/Affinity group discounts in order to attract corporate clients which could potentially boost the net revenue of the company.

An in-depth analysis was performed on the filtered data on each of the marketing channels in order to determine average claims filed by customers on-boarded by the particular channel of sales. The data suggests that the Agent attracted customers had the lowest claims on average (2.38 per customer) while the customers who signed up using the website had the highest claims on average (2.41 per customer). Based upon this information, the company should look to prioritize agent marketing over any other sales channel. This could ensure that the company wins over clients who are less likely to file more claims which in turn will enhance revenues. 

The company should offer special discounts exclusively to customers who have been with them for over 3 years. This is based on the fact that the customers with active policies in between 40 to 55 months turned out to be highly profitable in terms of revenue generated. The company's highly profitable clients had their policies active for an average of 46.42 months. The special discounts on customer loyalty would help the company with the client retention in turn leading to increased revenues.

Luxury SUVs, Luxury Cars, and Sports Cars generated lifetime revenues of $6,381, $5,669 and 2,158 on average respectively while the typical two-door and four-door cars netted $269 and $271 in revenues on average respectively. On the basis of this analysis, the company should work to diversity in order to attract clients owning Luxury and Sports cars. The company should offer discounts and lower premiums on these cars so as to become the client’s number one choice in this niche. This would help the company see increased revenues in the long term. 

The average Customer Lifetime Value of customers on the Basic, Extended and Premium coverage came out to be $326, $1,538 and, $2,879 respectively. The company should encourage more and more customers to opt for the Extended and Premium coverage by incentivizing those plans. 

Of all the customers generating more than average revenue for the company, the Suburban and Urban Women helped the company make $2,531,073 and $2,011,879 respectively as compared to their male counterparts ($552,967 and $ 815,521 respectively). The company had to pay a grand total of $1,213,122 to their male clients in claims while they had to spend a far lesser amount of $758,704 covering up their Female customer claims. Based upon these facts the company should work to execute highly targeted campaigns to attract Suburban and Urban women. The company could also offer lower premiums to females from these demographics which in turn could skyrocket the company’s revenues. 















